Fusing Filters with Integer Linear Programming

### 2014-05-15 2014betterfusionforfilters/section/03-SizeInference.tex
Wording. Replaced ref to normalise with normalize2

A size may either be a variable $k$, or a cross product of two sizes. We use the latter to represent the result size of the @cross@ operator discussed in the previous section. Constraints maybe either be trivially $true$, an equality $\tau = \tau$, or a conjunction of two constraints $C \wedge C$. We refer to the trivially true and equality constraints as \emph{atomic constraints}. Size schemes relate the sizes of each input and output array. For example, the size scheme for the @normalize@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Note that the constraints are generated in a specific form, to make the constraint solving process easy. For each array variable in the program we generate a new size variable, like size $k_{zs}$ for array variable $zs$. These new size variables always appear on the \emph{left} of atomic equality constraints. For each array binding we may also introduce unification or rigid variables, and these appear on the \emph{right} of atomic equality constraints.

For example, the final environment and constraints generated for the @normalize@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

During constraint solving, all unification variables mentioned in the environment can have other sizes substituted for them. In contrast, the rigid variables marked by the $\exists$ symbol cannot. For example, if we consider the constraints for @normalize@ mentioned before:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately. This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}. In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
$$@normalize@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
%The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
%\gabi{gts is no combinator, so can't be a transducer
The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
\begin{code}
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ...
\end{code}

>>>>

A size may either be a variable $k$ or a cross product of two sizes. We use the latter to represent the result size of the @cross@ operator discussed in the previous section. Constraints maybe either be trivially $true$, an equality $k = \tau$, or a conjunction of two constraints $C \wedge C$. We refer to the trivially true and equality constraints as \emph{atomic constraints}. Size schemes relate the sizes of each input and output array. For example, the size scheme for the @normalize2@ example from Figure~\ref{f:normalize2-clusterings} is as follows:
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Note that the constraints are generated in a specific form, to make the constraint solving process easy. For each array variable in the program, we generate a new size variable, like size $k_{zs}$ for array variable $zs$. These new size variables always appear on the \emph{left} of atomic equality constraints. For each array binding we may also introduce unification or rigid variables, and these appear on the \emph{right} of atomic equality constraints.

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

During constraint solving, all unification variables mentioned in the environment can have other sizes substituted for them. In contrast, the rigid variables marked by the $\exists$ symbol cannot. For example, if we consider the constraints for @normalize2@ mentioned before:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

To produce the final size scheme we lookup the sizes of the input and output variables of the original function from the solved constraints, and generalize appropriately. This process is determined by the top-level rule in Figure~\ref{f:ConstraintGeneration}. In this case no rigid size variables appear in the result, so we can universally quantify all size variables.
$$@normalize2@ ~:_s \forall k. (xs : k) \to (ys_1 : k,~ ys_2 : k)
$$

@@

Looking back at the @normalize2@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
%The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
%\gabi{gts is no combinator, so can't be a transducer
The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
\begin{code}
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ...
\end{code}
